Solution of thermodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2016-05-12 pituba/article/article.tex
news

% Review
\documentclass[authoryear,preprint,review,12pt]{elsarticle}
%\documentclass[final,authoryear,5p,times,twocolumn,11pt]{elsarticle}

@@

\begin{abstract}
This work presents a derivation of the classical material balance equation starting from black-oil equations in integral form. The existing link between both approaches is examined in detail, including the necessary step of averaging PVT properties. An interpretation for pressure in material balance calculations is provided and the choice of a reference depth (datum) is discussed. The material balance equation is formulated for a varying bubble-point pressure and sufficient conditions for solution uniqueness are presented. A scheme for numerical solution is also proposed.
\end{abstract}

@@

\section{Introduction}
The so called material balance equation is widely employed in reservoir engineering and has become a classical tool in oil reservoir management. It consists of an algebraic equation originally presented in \cite{schilthuis1936active} and derived from a zero dimensional model for estimating reservoir pressure from mass balance. \cite{schilthuis1936active} introduced the hypothesis of thermodynamic equilibrium occurring not in the entire reservoir volume, but in the regions of high permeability where most of the production would come from and where pressure would present a similar behavior. Such regions were said to contain an \textit{active oil}.

@@

The term black-oil refers to specific hydrocarbon compositions whose phase equilibrium for isothermal changes in pressure may be described by a two pseudo-component system. For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and we may define this composition as a pseudo gas component. A pseudo oil component may then be defined by subtracting the gas component from oil phase composition. After defining the two pseudo components, the number of transport equations needed for hydrocarbon species is reduced from usually a large number to only two.

@@

\subsection{Black-Oil equations}

As already mentioned, black-oil is a designation to a simplified treatment of hydrocarbon composition and phase equilibrium. Hydrocarbon species are grouped into two pseudo-components called oil and gas, being a pseudo-component one that consists of more than one chemical molecule.

Oil component, also called stock tank oil, is defined as the hydrocarbon composition in liquid phase for standard surface pressure and temperature. Similarly, gas component is the hydrocarbon composition in gaseous phase for the same surface conditions. All species are generally present in both phases although the equilibrium composition is shifted to liquid phase (or gaseous phase) for components of greater (lower) molecular mass.

Reservoir fluids are then composed of the two pseudo-components (oil and gas) plus the water component. There can exist three fluid phases: oil, gas and liquid water. Oil phase may be composed of oil and gas components, gas phase may be composed uniquely by the gas component and, water phase, uniquely by water component. The only allowed mass transfer is of gas component between oil and gas phases.

@@

A common simplification in black-oil formulations is to restrict mass transport to macroscopic advection. No diffusive or dispersive term is included in the mass balance equations. This simplification must be looked with care when gradients of species concentration are important. For instance, when gas is being dissolved in oil phase in a reservoir under repressurization, this condition is not generally true.

@@

\begin{equation}\label{eq: exdarcy}
\mathbf{u}_\alpha = -\frac{k_{r,\alpha} \mathbf{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right) \, , \quad \alpha=o,w,g \, .
\end{equation}

@@

Another simplification is now made assuming thermal equilibrium in the entire reservoir volume:
\begin{equation}
T_w = T_o = T_g = T_{rock} = T
\end{equation}
and
\begin{equation}
\nabla T \equiv 0 \, .
\end{equation}

@@

Porosity ($\phi$) is assumed a function of pore pressure ($p_P$). Since three phases are present in general, such quantity is defined by some relation of type \citep{kim2011rigorous}:
\begin{equation}
p_P=p_P\left(p_o,p_w,p_g,S_o,S_w,S_g\right) \, .
\end{equation}

@@

We start assuming mechanical equilibrium, i.e., no bulk motion:
\begin{equation}
\mathbf{u}_\alpha = -\frac{k_{r,\alpha} \mathbf{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right)\equiv 0 \, .
\end{equation}
This condition implies that pressure gradients must equal gravity force per unit volume:
\begin{equation}\label{eq: p=rhogh}
\nabla p_\alpha = \rho_\alpha g\nabla z
\end{equation}
and, consequently, pressure is a function of depth only:
\begin{equation}\label{eq: p=rhogh}
\int_{p_\alpha^0}^{p_\alpha} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'},T\right)} = g\left(z - z_0\right) \, .
\end{equation}

@@

We proceed omitting function arguments:
\begin{equation}
\bar{S_\alpha}\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}S_\alpha \phi dV \, ,
\end{equation}

\begin{equation}
\bar{\rho_\alpha}\bar{S_\alpha}\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}\rho_\alpha S_\alpha \phi dV \, ,
\end{equation}

\begin{equation}
\bar{\rho_\alpha}\bar{Y_{\alpha,k}}\bar{S_\alpha}\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}\rho_\alpha Y_{\alpha,k} S_\alpha \phi dV \, .
\end{equation}

@@

\begin{itemize}
\item water component:
\begin{equation}\label{eq: Sw3}
\frac{\partial}{\partial t} \left( \bar{\rho_w}\bar{S_w}\bar{\phi} \Omega \right) = \rho_w^{ST}Q_w^{ST}
\end{equation}

\item oil component:
\begin{equation}\label{eq: So3}
\frac{\partial}{\partial t} \left( \bar{\rho_o}\bar{Y_o}\bar{S_o}\bar{\phi} \Omega \right) = \rho_o^{ST}Q_o^{ST}
\end{equation}

\item gas component:
\begin{equation}\label{eq: Sg3}
\frac{\partial}{\partial t} \left[\left( \bar{\rho_o}\bar{Y_g} \bar{S_o} + \bar{\rho_g} \bar{S_g} \right) \bar{\phi} \Omega\right] = \rho_g^{ST}Q_g^{ST} \, .
\end{equation}
\end{itemize}

@@

Material balance equation is widely employed to stablish a relationship between production (and injection) and reservoir pressure. It turns out, though, that pressure is different for different locations, mostly for different depths, and also for different phases.
What is then the significance of having a single value of pressure for the entire reservoir volume?

For answering this question, we examine the two roles played by pressure in material balance equation: computation of averaged phase density and phase equilibrium. We shall see that for the purpose of computing the averaged phase density, we may define some reference depth and use the corresponding local density as an approximation to the averaged density without introducing any important error. Pressure increases with depth according to the hydrostatic gradient. Consequently, density also increases but by a negligible amount.

@@

A single equation for pressure may be obtained by substituting Eqs.~\eqref{eq: So5}, \eqref{eq: Sg5} and \eqref{eq: Sw5} into \eqref{eq: S15}. The resulting equation allows pressure to be calculated from given production ($F_p$) and injection ($F_i$) volumes:
\begin{equation}\label{eq: MBE}
F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p1}-V_{p2}}_{\text{porous volume contraction } (C)}
\end{equation}
or
\begin{equation}\label{eq: MBE_short}
F_p - F_i= E + C \, ,
\end{equation}

@@

\section{Solution of Material balance equation}
A numerical solution of Eq.~\eqref{eq: MBE_short} is presented here. We shall start defining a residual function ($\varepsilon$) for which the pressure solving Eq.~\eqref{eq: MBE_short} is a root:
\begin{equation}\label{eq: residual}
\varepsilon(p_2) = \frac{E+F_i+C-F_p}{V_{p1}} \, .
\end{equation}
We proceed showing that $\varepsilon$ is a monotonic function provided some conditions on fluid properties are observed and that, as a consequence, only one root of $\varepsilon$ exists at most. Equivalently, we may say that only one solution exists for material balance equation.

@@

Residual function clearly satisfies the first two affirmatives because fluid properties are continuous everywhere and are also differentiable everywhere except for the first order phase change at bubble-point pressure.
The last affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels corresponding each one to a different phase. Each phase parcel is composed of two factors: current volume \textit{in-situ} and volume derivative with respect to pressure:
\begin{equation}
\begin{split}
&V_{p1}\frac{d\varepsilon}{dp}=\underbrace{\left(N-N_p\right) \left(\frac{dB_o}{dp} - B_g \frac{dR_s}{dp} \right)}_\text{oil phase}\\
&+\underbrace{\left[N R_{s1}+mN\frac{B_{o1}}{B_{g1}}+G_i-\left(N-N_p\right)R_{s2}-G_p\right] \frac{dB_g}{dp}}_\text{gas phase}\\
&+\underbrace{\left(W + W_i+W_e-W_p\right)\frac{dB_w}{dp}}_\text{water phase}\\
&\underbrace{-\frac{dV_p}{dp}}_\text{porous volume}\, .
\end{split}
\end{equation}

\textit{In-situ} volumes (e.g. $N-N_p$ for oil phase) are non-negative values. Pressure derivatives of fluid volumes are all negative values (as fluid pressure increases, fluid volume contracts) and pressure derivative of rock porous volume is a positive value (as fluid pressure increases, porous volume expands).

@@

\subsection{Solution Algorithm}
Several methods exist to find roots of nonlinear continuous functions \citep{hamming1987numerical}. One possible choice is Newton's method, which linearizes the function using either its gradient or a secant line. The secant line approach introduces an artificial smoothness to the gradient direction near continuous but non-differentiable points, for what it is the approach employed here to deal with the bubble-point pressure.

@@

\begin{table*}
\small
\centering
\begin{tabular}{ll}
\toprule
\multicolumn{2}{l}{\textit{Material balance equation}}\\
\midrule
oil & $S_{o2} = \left(N-N_p\right) B_{o2}/V_{p2}$ \label{eq: oiltab}
\\
water & $S_{w2} = \left(W+W_i+W_e-W_p\right) B_{w2}/V_{p2}$\\
gas & $S_{g2}=\left\lbrace N\left[\left(R_{s1}-R_{s2}\right)+m B_{o1}/B_{g1}\right]+N_p R_{s2}+G_i-G_p\right\rbrace Bg_2/V_{p2}$ \\
bubble-point pressure & $R_{s2}\left(P_b\right) = \left[N\left(m B_{1}/B_{g1} +R_{s1}\right)+G_i-G_p\right]/\left(N-N_p\right)$ \\
pressure & $F_p - F_i= N E_o + mN E_g + W E_w +C$ \\
\midrule
\multicolumn{2}{l}{\textit{Definition of terms in pressure equation}} \\
\midrule
fluid production & $F_p=N_p \left[B_{o2} + \left(G_p/N_p -R_{s2}\right)B_{g2}\right]+W_p B_{w2}$ \\
fluid injection and influx & $F_i=\left(W_i +W_e\right) B_{w2}+G_{i}B_{g2}$ \\
oil phase expansion & $E_o=\left(B_{o2}-B_{o1}\right) - \left(R_{so2}-R_{so1}\right)B_{g2}$ \\
gas phase expansion & $E_g = B_{o1}\left(B_{g2}/B_{g1}-1\right)$ \\
water volume expansion & $E_w = B_{w2}-B_{w1}$\\
porous volume contraction & $C=V_{p2}- V_{p1}$\\ \bottomrule
\end{tabular}
\caption{Complete set of material balance equations.}
\label{tab: MBE}
\end{table*}

>>>>

% Review
%\documentclass[authoryear,preprint,review,10pt]{elsarticle}
\documentclass[final,authoryear,5p,times,twocolumn,11pt]{elsarticle}

@@

\begin{abstract}
This work presents a derivation of the classical material balance equation starting from black-oil equations in integral form. The existing link between both approaches is examined, including the step of averaging PVT properties. An interpretation for pressure in material balance calculations is provided and the choice of a reference depth (datum) is discussed. The material balance equation is formulated for a varying bubble-point pressure and sufficient conditions for solution uniqueness are presented. A scheme for numerical solution is also proposed for the case when pressure is the unknown quantity.
\end{abstract}

@@

\section{Introduction}
The so called material balance equation is widely employed in reservoir engineering and has become a classical tool in oil reservoir management. It consists of an algebraic equation originally presented in \cite{schilthuis1936active} and derived from a zero dimensional model for estimating reservoir pressure from mass balance. \cite{schilthuis1936active} introduced the hypothesis of thermodynamic equilibrium occurring not in the entire reservoir volume, but in the regions of high permeability where most of the production would come from and where pressure would present a similar trend. Such regions were said to contain an \textit{active oil}.

@@

The term black-oil refers to specific hydrocarbon compositions whose phase equilibrium for isothermal changes in pressure can be described by a two pseudo-component system. For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and we may define this composition as a pseudo gas component. A pseudo oil component may then be defined by subtracting the gas component from oil phase composition. After defining the two pseudo components, the number of transport equations needed for hydrocarbon species is reduced from usually a large number to only two.

@@

\subsection{Black-oil equations}

As mentioned, black-oil is a designation to a simplified treatment of hydrocarbon composition and phase equilibrium. Hydrocarbon species are grouped into two pseudo-components called oil and gas, being a pseudo-component one that consists of more than one chemical molecule.

Oil component, also called stock tank oil, is defined as the hydrocarbon composition in liquid phase for standard surface pressure and temperature. Similarly, gas component is the hydrocarbon composition in gaseous phase for the same surface conditions. Reservoir fluids are then composed of the two pseudo-components (oil and gas) plus the water component. There can exist three fluid phases: oil, gas and liquid water. Oil phase may be composed of oil and gas components, gas phase may be composed uniquely by the gas component and, water phase, uniquely by water component. The only allowed mass transfer is of gas component between oil and gas phases.

@@

A common simplification in black-oil formulations is to restrict mass transport to advection. No diffusive term is included in the mass balance equations. This simplification must be looked with care when gradients of species concentration are important. For instance, when gas is being dissolved in oil phase in a reservoir under repressurization, this condition is not generally true.

@@

\begin{equation}\label{eq: exdarcy}
\mathbf{u}_\alpha = -\frac{k_{r,\alpha} \mathbf{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha \mathbf{g}\right) \, , \quad \alpha=o,w,g \, .
\end{equation}

@@

Another simplification is now made assuming thermal equilibrium in the entire reservoir volume:
\begin{equation}
T_w = T_o = T_g = T_{rock} = T\, , 
\end{equation}
and
\begin{equation}
dT \equiv 0 \, .
\end{equation}

@@

Porosity ($\phi$) is assumed a function of pore pressure ($p_P$). Since three phases are present in general, such quantity is defined by some relationship as follows \citep{kim2011rigorous}:
\begin{equation}
p_P=p_P\left(p_o,p_w,p_g,S_o,S_w,S_g\right) \, .
\end{equation}

@@

We start assuming mechanical equilibrium, i.e., no bulk motion:
\begin{equation}
\mathbf{u}_\alpha = -\frac{k_{r,\alpha} \mathbf{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha \mathbf{g}\right)\equiv 0 \, .
\end{equation}
This condition implies that pressure gradients must equal gravity force per unit volume:
\begin{equation}\label{eq: p=rhogh}
\nabla p_\alpha = \rho_\alpha \mathbf{g} \, .
\end{equation}
Consequently, pressure is a function of depth only:
\begin{equation}\label{eq: p=rhogh}
\int_{p_\alpha^0}^{p_\alpha} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'},T\right)} = g\left(z - z_0\right) \, .
\end{equation}

@@

We proceed omitting function arguments:
\begin{equation}
\bar{S}_\alpha\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}S_\alpha \phi dV \, ,
\end{equation}

\begin{equation}
\bar{\rho_\alpha}\bar{S}_\alpha\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}\rho_\alpha S_\alpha \phi dV \, ,
\end{equation}

\begin{equation}
\bar{\rho}_\alpha\bar{Y}_{\alpha,k}\bar{S}_\alpha\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}\rho_\alpha Y_{\alpha,k} S_\alpha \phi dV \, .
\end{equation}

@@

\begin{itemize}
\item water component:
\begin{equation}\label{eq: Sw3}
\frac{\partial}{\partial t} \left( \bar{\rho}_w\bar{S}_w\bar{\phi} \Omega \right) = \rho_w^{ST}Q_w^{ST}
\end{equation}

\item oil component:
\begin{equation}\label{eq: So3}
\frac{\partial}{\partial t} \left( \bar{\rho}_o\bar{Y}_o\bar{S}_o\bar{\phi} \Omega \right) = \rho_o^{ST}Q_o^{ST}
\end{equation}

\item gas component:
\begin{equation}\label{eq: Sg3}
\frac{\partial}{\partial t} \left[\left( \bar{\rho}_o\bar{Y}_g \bar{S}_o + \bar{\rho}_g \bar{S}_g \right) \bar{\phi} \Omega\right] = \rho_g^{ST}Q_g^{ST} \, .
\end{equation}
\end{itemize}

@@

Material balance equation is widely employed to stablish a relationship between production (and injection) and reservoir pressure. It turns out, though, that pressure is different for different locations and also for different phases.

What is then the significance of having a single value of pressure for the entire reservoir volume?

For answering this question, we examine the two roles played by pressure in material balance equation: computation of phase density and phase equilibrium.

We shall see that for the purpose of computing the averaged phase density, we may define some reference depth and use the corresponding local density as an approximation to the averaged density without introducing any important error. Pressure increases with depth according to the hydrostatic gradient. Consequently, density also increases but by a negligible amount.

@@

A single equation for pressure may be obtained by substituting Eqs.~\eqref{eq: So5}, \eqref{eq: Sg5} and \eqref{eq: Sw5} into \eqref{eq: S15}. The resulting equation allows pressure to be calculated from given production ($F_p$) and injection ($F_i$) volumes:
\begin{equation}\label{eq: MBE}
F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p2}-V_{p1}}_{\text{porous volume change } (\Delta V_p)}
\end{equation}
or
\begin{equation}\label{eq: MBE_short}
F_p - F_i= E - \Delta V_p \, ,
\end{equation}

@@

\section{Solution of material balance equation}
A numerical solution of Eq.~\eqref{eq: MBE_short} is presented here. We start defining a residual function ($\varepsilon$) for which the pressure solving Eq.~\eqref{eq: MBE_short} is a root:
\begin{equation}\label{eq: residual}
\varepsilon(p_2) = \frac{E+F_i-\Delta V_p-F_p}{V_{p1}} \, .
\end{equation}
We proceed showing that $\varepsilon$ is a monotonic function provided some conditions on fluid properties are observed and that, as a consequence, only one root of $\varepsilon$ exists at most. Equivalently, we may say that only one solution exists for material balance equation.

@@

Residual function clearly satisfies the first two affirmatives because fluid properties are continuous everywhere and are also differentiable everywhere except for the first order phase change at bubble-point pressure.
The last affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels corresponding each one to a different phase. Each phase parcel is composed of two factors: current volume \textit{in-situ} and volume derivative with respect to pressure:
\begin{equation}
\begin{split}
&V_{p1}\frac{d\varepsilon}{dp}=\underbrace{\left(N-N_p\right) \left(\frac{dB_o}{dp} - B_{g2} \frac{dR_s}{dp} \right)}_\text{oil phase}\\
&+\underbrace{\left[N R_{s1}+mN\frac{B_{o1}}{B_{g1}}+G_i-\left(N-N_p\right)R_{s2}-G_p\right] \frac{dB_g}{dp}}_\text{gas phase}\\
&+\underbrace{\left(W + W_i+W_e-W_p\right)\frac{dB_w}{dp}}_\text{water phase}\\
&\underbrace{-\frac{dV_p}{dp}}_\text{porous volume}\, .
\end{split}
\end{equation}

Remaining volumes in surface condition (e.g. $N-N_p$ for oil phase) are non-negative values. Pressure derivatives of fluid volumes are all negative values (as fluid pressure increases, fluid volume contracts) and pressure derivative of rock porous volume is a positive value (as fluid pressure increases, porous volume expands).

@@

\subsection{Solution algorithm}
Several methods exist to find roots of nonlinear continuous functions \citep{hamming1987numerical}. One possible choice is Newton's method, which linearizes the function using either its gradient or a secant line. The secant line approach introduces an artificial smoothness to the gradient direction near continuous but non-differentiable points, for what it is the approach employed here to deal with the bubble-point pressure.

@@

\begin{table*}
\small
\centering
\begin{tabular}{ll}
\toprule
\multicolumn{2}{l}{\textit{Material balance equation}}\\
\midrule
oil & $S_{o2} = \left(N-N_p\right) B_{o2}/V_{p2}$ \label{eq: oiltab}
\\
water & $S_{w2} = \left(W+W_i+W_e-W_p\right) B_{w2}/V_{p2}$\\
gas & $S_{g2}=\left\lbrace N\left[\left(R_{s1}-R_{s2}\right)+m B_{o1}/B_{g1}\right]+N_p R_{s2}+G_i-G_p\right\rbrace Bg_2/V_{p2}$ \\
bubble-point pressure & $R_{s2}\left(P_b\right) = \left[N\left(m B_{1}/B_{g1} +R_{s1}\right)+G_i-G_p\right]/\left(N-N_p\right)$ \\
pressure & $F_p - F_i= N E_o + mN E_g + W E_w - \Delta V_p$ \\
\midrule
\multicolumn{2}{l}{\textit{Definition of terms in pressure equation}} \\
\midrule
fluid production & $F_p=N_p \left[B_{o2} + \left(G_p/N_p -R_{s2}\right)B_{g2}\right]+W_p B_{w2}$ \\
fluid injection and influx & $F_i=\left(W_i +W_e\right) B_{w2}+G_{i}B_{g2}$ \\
oil phase expansion & $E_o=\left(B_{o2}-B_{o1}\right) - \left(R_{so2}-R_{so1}\right)B_{g2}$ \\
gas phase expansion & $E_g = B_{o1}\left(B_{g2}/B_{g1}-1\right)$ \\
water volume expansion & $E_w = B_{w2}-B_{w1}$\\
porous volume contraction & $\Delta V_p=V_{p2}- V_{p1}$\\ \bottomrule
\end{tabular}
\caption{Complete set of material balance equations.}
\label{tab: MBE}
\end{table*}
